Spatial Reasoning and Statistical Graphics

### 2014-11-06 Paper/Draft1/TestingVisualAptitude.Rnw
Fix PCA discussion to emphasize lineups = figure classification

options(replace.assign=TRUE,width=30,scipen=3)

@@

Figure classification is strongly related to lineups (PC2, PC3), and as in the four-component PCA, figure classification is strongly represented in both principal components. 
Performance on the visual search task is also related to lineup performance (PC3), and is more strongly represented in PC3 in the five-component PCA than in the four-componet PCA. 
These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

While lineups do span a separate dimension, but the PCA also suggests that they are most closely related to the figure classification task, and least related to the visual searching task.

The additional dimension, PC4 in the five-component PCA, separates lineups from the figure classification task and incorporates the relatively weak correlation between performance on the card rotation task and lineup performance; this dimension accounts for \Sexpr{round(summary(pca)$importance[2,4]*100,1)}\% of the variance. 
PC5 is almost identical to PC4 in the four-component PCA; as before, it does not account for a significant portion of the variance in the data.

This emphasizes the underpinnings of lineups: they are a visual medium, but they ultimately are a classification task, presented in a graphical manner. Using lineups as a proxy for statistical significance tests is similar to using a classifier on pictoral data: while the data is presented ``graphically", the participant is actually classifying the data based on underlying summary statistics.

\subsection{Linear model of demographic factors}
\newtext{Note that all of the demographic variables in the survey are highy correlated, for example there is a high correlation between STEM majors and taking calculus.% -- only eight students are either not STEM, but have taken calculaus 1 or are STEM but have not taken the class. 
Similaryl, the correlation  between having taken a statistics class and having been involved in mathematics or statistics research is high. Only one student is doing research who has (not yet ?) taken a statistics course. A principal component of the five math/stats questions splits the variables into two main areas: the first principal component is an average of math skills, calculus 1 and STEM, while the second principal component is an average of having taken a statistics class and doing research. We therefore decided to use sums of these variables to come up with a separate math and a stats score. Note, that the correlation between the math and the stats score is almost zero. }
<<linearmodel, results='hide', echo=FALSE>>=
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
ans.model <- data.frame(ans.summary, data.frame(pca2$x))
# exclude all variables that are not supposed to go into the linear model
# those are the unrotated tests; instead we have the PCs
excl <- c("id", "card_rot", "fig_class", "folding", "vis_search", "vidgame_hrs", "major1", "major2", "minor1", "minor2", "learning_disability", "colorblind", "epilepsy", "normal_vision")

@@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
\newtext{We fit a linear model of lineup scores in the thus modified demographic variables and the test scores from the visuo-spatial tests. Using a step-wise backwards selection based on AIC (Akaike Information Criterion - do we need a reference?) we select the best model. The result is shown in table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as an innate visual ability, while the MATH score is a matter of both ability and training. 
}
<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
\comment{
results need to go here
}

>>>>

options(replace.assign=TRUE,scipen=3)

@@

Figure classification is strongly related to lineups, and as in the four-component PCA, figure classification is strongly represented in the first two principal components. While lineups do span a separate dimension, the PCA suggests that they are most closely related to the figure classification task, and least related to the visual searching task.

% 
% The additional dimension, PC4 in the five-component PCA, separates lineups from the figure classification task and incorporates the relatively weak correlation between performance on the card rotation task and lineup performance; this dimension accounts for \Sexpr{round(summary(pca)$importance[2,4]*100,1)}\% of the variance. 
% PC5 is almost identical to PC4 in the four-component PCA; as before, it does not account for a significant portion of the variance in the data.
%

This emphasizes the underpinnings of lineups: the test utilizes a visual medium, but is ultimately are a classification task presented in a graphical manner. Using lineups as a proxy for statistical significance tests is similar to using a classifier on pictoral data: while the data is presented ``graphically", the participant is actually classifying the data based on underlying summary statistics.

\subsection{Linear model of demographic factors}
\newtext{Note that all of the demographic variables in the survey are highy correlated, for example there is a high correlation between STEM majors and taking calculus.% -- only eight students are either not STEM, but have taken calculaus 1 or are STEM but have not taken the class. 
Similarly, the correlation  between having taken a statistics class and having been involved in mathematics or statistics research is high. Only one student is doing research who has (not yet ?) taken a statistics course. A principal component of the five math/stats questions splits the variables into two main areas: the first principal component is an average of math skills, calculus 1 and STEM, while the second principal component is an average of having taken a statistics class and doing research. We therefore decided to use sums of these variables to come up with a separate math and a stats score. Note, that the correlation between the math and the stats score is almost zero. }
<<linearmodel, results='hide', echo=FALSE>>=
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
ans.model <- data.frame(ans.summary, data.frame(pca2$x))
# exclude all variables that are not supposed to go into the linear model
# those are the unrotated tests; instead we have the PCs
excl <- c("id", "card_rot", "fig_class", "folding", "vis_search", "vidgame_hrs", "major1", "major2", "minor1", "minor2", "learning_disability", "colorblind", "epilepsy", "normal_vision")

@@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
\newtext{We fit a linear model of lineup scores in the thus modified demographic variables and the test scores from the visuo-spatial tests, selecting the best model using AIC and stepwise backwards selection. The result is shown in table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as a measure of innate visual or intellectual ability, while the MATH score is a matter of both ability and training. 
}
<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
\comment{
results need to go here
}
